File-based localization of numerical perturbations in data analysis pipelines

### 2020-03-24 paper.tex
Some comments on results

\usepackage{subcaption}
\usepackage{xspace}
\usepackage{stmaryrd} % for llbracket and rrbracket
\usepackage{amsmath}
\usepackage{ulem}
\usepackage{pifont}
\DeclareMathOperator*{\argmin}{argmin}

@@

In this experiment, we analyse the numerical reproducibility of computational pipelines 
and identify the origin of differences across the operating systems. 
Two types of differences can occur in the subjects due to the differences
in the operating systems. One is between-OS differences caused by the
operating system library updates and the other type, within-OS differences
occur as a result of the pseudo-random processes used in the pipelines.

In particular, Spot tool is tested on the neuroimaging applications which are 
predominantly using mathematical libraries. Therefore, we expect to find 
differences as a result of changing the mathematical functions between operating system libraries.

This section describes datasets and pipelines used for the analyses, and the way of data processing.

@@

We used our tool to evaluate the reproducibility of the pipelines from the Human Connectome 
Project (\href{https://www.humanconnectome.org}{HCP}).
The HCP initiative is an effort to acquire and analyse 
brain connectivity data from 1200 healthy adults.
It enables the neuroscience 
research community to discover relationships between brain circuits and 
individual behaviors. This helps to understand a wide range of brain disorders.
The HCP project provides database services (ConnectomeDB) for storing and 
sharing primary and processed data freely, and data analysis pipelines that 
are available under an open-source license.

The HCP developed a set of pre-processing pipelines to process structural,
functional, and diffusion MRI data acquired in the project. We focus on HCP
pre-processing pipelines for structural data, particularly PreFreeSurfer
and FreeSurfer.

According to~\cite{glasser2013}, the sequence of the 
sub-pipelines in PreFreeSurfer includes 
(1) Distortion Correction (DC), 
(2) Anatomical Average (AAve), 
(3) ACPC (Anterior Commissure, Posterior Commissure) Alignment (ACPC-A), 
(4) Brain Extraction (BExt), 
(5) Bias Field Correction (BFC), 
(6) Atlas-Registration (AR).

Also, the sequence of the main sub-pipelines in FreeSurfer includes 
(1) Downsampling images, 
(2) T1w image registration, 
(3) Placement of surface images, 
(4) Folding-based surface registration.
A detailed description of the analyses done in these
pipelines is available in~\cite{glasser2013}.

Furthermore, the average time per subject taken for PreFreeSurfer processing is $\approx$ 2.2 hours, 
and FreeSurfer processing is $\approx$ 8.3 hours (20 runs), 
and the average output file size of 20 subjects in PreFreeSurfer is $\approx$ 2.7 GB, 
and FreeSurfer is $\approx$ 4.1 GB.

HCP data collected from different types of imaging techniques, including 
structural imaging (sMRI), functional imaging (fMRI), and diffusion imaging (dMRI).
The structural images include T1-weighted (T1w) and T2-weighted (T2w) images, which 
help in the diagnosis of brain injury.
The images are indexed with a suffix if several scans of the same modality were acquired.
For example, some data may include two T1w images with low and high resolutions.
The functional images, including task-based fMRI and resting-state scans, 
enable the measurement of functional activations within brain areas. 
Diffusion imaging is another kind of MRI technique, which measures 
the anatomical connectivity between regions.
See~\cite{van2013wu} for a discussion of these various modalities.

In this paper, we randomly selected 20 unprocessed subjects 
containing 3-tesla type of MRI scanners from the HCP data release S500 
available in \href{https://db.humanconnectome.org}{the ConnectomDB repository}.

@@

We executed the pipelines using Docker containers to simplify the 
deployment of different operating system versions on execution 
platforms. 
The Docker images were built for the HCP pre-processing 
pipelines v3.19.0 (PreFreeSurfer and FreeSurfer) in 
CentOS 6.9 (Final) and CentOS 7.4 (Core). 
CentOS is a widely used operating system among the Neuroscience community because of 
the appropriate community support, network functions, safety, portability, and openness
~\cite{hanke2011neuroscience}.
These containers were installed with all the necessary software and libraries 
required to run the HCP pipelines including FSL (version 5.0.6), 
FreeSurfer (version 5.3.0-HCP, CentOS4 build), and Connectome Workbench (version 1.0).
Docker images are available on 
\href{https://hub.docker.com/r/bigdatalabteam/hcp-prefreesurfer/}{DockerHub}
for reuse.

The data were processed with PreFreeSurfer and FreeSurfer. 
First, PreFreeSurfer was utilized to analyse the unprocessed data, and then its 
results were used as the input of FreeSurfer.
Each subject was processed twice on the same operating system to detect
variability coming from pseudo-random operations. 
Within-OS variability may also be coming from other factors than pseudo-randomness. 
For instance, the same files may have different metadata such as timestamps, 
file paths, and other execution-specific information. 
To ensure that the differences are not related to the metadata, 
we used specific functions from FreeSurfer tool including 
\emph{mri\_diff}, \emph{mris\_diff}, and \emph{lta\_diff}.
These functions compare files based on the image dimensions, 
voxel intensities, the number of vertices, and geometry information
(see~\cite{fischl2012freesurfer} for more details). 
In addition, we used the Dice coefficient metric to evaluate the similarity of 
the image segmentation results, which is defined by
\[DICE=\frac{2(|X|\cap|Y|)}{|X|+|Y|}\]
The Dice coefficient ranges from 0, indicating X and Y have no voxels in common, to 1, if 
two images are identical.

Finally, to cluster the subjects, the threshold value is set to zero in the clustering method. 
Also, we used the nearest neighbor algorithm to calculate the distance between clusters.

@@

We did not observe any within-OS difference in PreFreeSurfer, however, we
identified two FreeSurfer processes leading to within-OS differences due
to the use of pseudo-random numbers: image registration with
\emph{mri\_segreg}, and cortical surface curvature
estimations with \emph{mris\_curvature}. To address this issue, we fixed
the seed used by the pseudo-random number generator, through defining 
option ``--seed'' in the command-line of the processes.
Our analysis can focus on between-OS differences.

@@

We identified four types of subjects with different PreFreeSurfer process
trees (Table~\ref{table:data-clusters}). Differences between subject types
are coming from different  numbers of T1 and T2 images in the raw data:
subjects of type 1 have 2 T1 images and 2 T2 images, while subjects of type
2 only have 1 T1 image and 1 T2 image. This leads to
different sets of output files. We verified that the process trees were
identical for all subjects of the same type, for both versions of
CentOS.

\begin{table}
\centering
\begin{threeparttable}
\caption{Clustering of 20 input subjects.}
\label{table:data-clusters}

\begin{tabular}{cccccc}
\toprule
       &                        &  \multicolumn{4}{c}{Available Modalities}    \\ 
\cmidrule(lr){3-6}       
Type   &   \makecell{Number of \\ Subjects}   &  T1w\_1          & T1w\_2      & T2w\_1          & T2w\_2        \\ \midrule
1      &               9                      &   \ding{51}      &   \ding{51} &   \ding{51}     &   \ding{51}   \\ 
2      &               8                      &   \ding{51}      &             &   \ding{51}     &               \\ 
3      &               1                      &   \ding{51}      &             &   \ding{51}     &   \ding{51}   \\ 
4      &               2                      &   \ding{51}      &   \ding{51} &   \ding{51}     &               \\ 
\bottomrule
\end{tabular}
\begin{tablenotes}
     \small
     \item *Type refers to the topology type of the process tree of subjects.
\end{tablenotes}
\end{threeparttable}
\end{table}

Figure~\ref{fig:pfs_freq} shows the frequency at which  processes create differences 
in the pipeline, and how differences are distributed among sub-pipelines.
Differences were observed in linear registration 
with FSL FLIRT (in ACPC-Alignment, Brain Extraction, Distortion Correction, 
Atlas Registration), in non-linear registration with FSL FNIRT (in Brain Extraction 
and Altas Registration), and in image warping with FSL \emph{new\_invwarp} (in Brain Extraction 
and Atlas Registration). Differences were also observed in image mean 
computations with FSL maths  (in Anatomical Average).

The results show two types of differences, between-process and between-subject.
Between-process differences occur when the same tool behaves differently in different process instances.
For example, 
FLIRT tool always produces files without differences in \emph{AAve} sub-pipeline but 
mostly creates different files in \emph{AAli} sub-pipeline.
Also, different frequencies show between-subjects variability.
Accordingly, the same process can be reproducible for some subjects but not for others. 
The similar findings in another study~\cite{bowring2019exploring} were reported 
the effect of between-subject differences on reproducibility.
% Between subject differences can be a result of different subject types. 
% For instance, the subjects contain two T1 images need more calculations 
% to produce an average image between them.

% make figure in two columns use {figure*}
\begin{figure*}
\centering
  \includegraphics[width=\textwidth]{images/pfs_heatmap.png}
  \caption{Frequency heatmap of processes in the tools that introduce differences among 20 subjects. 
           Each column represents the pipeline tool, and each row shows the sub-pipeline.
           Each cell may contain several instances of the process of a tool in a sub-pipeline.
           Annotations describe the frequency distribution of the instances as ``N [M] x Y" 
           in which N is the proportion of the process 
           that creates differences out of all M times that is occurred. 
           Also, times Y shows the number of processes with the same distribution.
           For example, the cell of FLIRT tool in the sub-pipeline AAve shows 10 instances of the process 
           with the frequency of 11 that introduce no differences.
           The range between red and green colors shows the highest and lowest 
           proportion of occurrences of the processes with differences.
           Also, grey cells indicate that the tool is not executed in the sub-pipeline.}
  \label{fig:pfs_freq}
\end{figure*}

Figure~\ref{fig:fnirt_result} 
shows checkerboard image from the results of ACPC-Alignment in PreFreeSurfer. 
Each rectangle includes the replaced voxels with different intensity values between images.
The misaligned regions visually represent significant differences in the alignment 
procedure.

%Alternating rectangular regions between two images, creating a characteristic 
%checkerboard-like pattern of varying magnitudes.

\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{images/t2w_alignment.png} 
    \caption{Differences between FNIRT results of one subject from PreFreeSurfer 
    (checkerboard image \emph{T2w\_acpc\_to\_MNI\_nonli.nii.gz}) (CentOS6 vs. 
    CentOS7).
    \href{https://github.com/ali4006/HCP-reproducibility-paper/blob/master/images/pfs_t2w_alignment.gif}
    {The link} to the animated version of alignment results shows the better visualization of differences.
} 
    \label{fig:fnirt_result}
\end{figure}

@@

We processed 20 subjects using FreeSurfer and 
Spot tool identified \emph{mris\_make\_surfaces} as the only process in 
the pipeline that creates differences in 10 subjects.
This process generates surface files for cortical and white matter surfaces 
(see~\cite{fischl2012freesurfer} for more detail). 
Also, FreeSurfer did not introduce any differences in 10 subjects.

Results show that the static builds of FreeSurfer in which uses the embedded libraries 
significantly avoids the creation of differences.
However, we found that between-OS differences occur when pre-processed data are different in FreeSurfer. 
FreeSurfer may not introduce differences, but it propagates and amplifies 
the current differences that appeared in PreFreeSurfer results across the operating systems.

Figure~\ref{fig:tissue_class} illustrates the sum of binarized differences 
from 20 subject results of the brain tissue segmentation across OSes.
This figure shows between-OS differences  
as a result of the propagation and amplification of differences in FreeSurfer. 
Differences are mostly localized at the interfaces between tissues.
In our dataset, we found subjects with min Dice value of 0.89 and max Dice  
value of 0.96.

Furthermore, the scatter plots of the Dice coefficients of the 44 segmented regions obtained 
from two operating systems are shown in Figure~\ref{fig:scatter_plot}.
This figure shows high Dice coefficients (more than 0.95) in large regions and very low 
Dice coefficients (below 0.5) in small structures particularly in the vessels and lateral ventricles.

\begin{figure}
%  \includegraphics{brain\_classification}
\centering
  \includegraphics[width=\columnwidth]{images/brain_segmentation_mni.png} 
  \caption{Heatmap binarized differences between brain segmentation results (file \emph{aseg.hires.mgz}) from 
          20 subjects in FreeSurfer (CentOS6 vs. CentOS7). 
          All binarized differences were resampled to the default MNI152 volume template.
          \href{https://github.com/ali4006/HCP-reproducibility-paper/blob/master/images/fs_brain_segmentation.gif}
          {The link} to the animated version of segmentation results of one subject presents more clear illustration of differences.} 
  \label{fig:tissue_class}
\end{figure}

@@

>>>>

\usepackage{subcaption}
\usepackage{xspace}
\usepackage{stmaryrd} % for llbracket and rrbracket
\usepackage{amsmath}
\usepackage{ulem}
\usepackage{pifont}

\hypersetup{
colorlinks=true,% hyperlinks will be coloured
linkcolor=blue,% hyperlink text will be blue
linkbordercolor=blue,% hyperlink border will be blue
}
\makeatletter
\Hy@AtBeginDocument{%
  \def\@pdfborder{0 0 1}% Overrides border definition set with colorlinks=true
  \def\@pdfborderstyle{/S/U/W 1}% Overrides border style set with colorlinks=true
                                % Hyperlink border style will be underline of width 1pt
}
\makeatother

\DeclareMathOperator*{\argmin}{argmin}

@@

We applied \toolname to find the origin of between-OS differences that
occur in many neuroimaging pipelines. In particular, we studied the minimal
pre-processing pipelines released by the Human Connectome Project
(\href{https://www.humanconnectome.org}{HCP}), a leading initiative in
neurosciences.

% In this experiment, we analyse the numerical reproducibility of computational pipelines 
% and identify the origin of differences across the operating systems. 
% Two types of differences can occur in the subjects due to the differences
% in the operating systems. One is between-OS differences caused by the
% operating system library updates and the other type, within-OS differences
% occur as a result of the pseudo-random processes used in the pipelines.

% In particular, Spot tool is tested on the neuroimaging applications which are 
% predominantly using mathematical libraries. Therefore, we expect to find 
% differences as a result of changing the mathematical functions between operating system libraries.

% This section describes datasets and pipelines used for the analyses, and the way of data processing.

@@

% We used our tool to evaluate the reproducibility of the pipelines from the Human Connectome 
% Project .
% The HCP initiative is an effort to acquire and analyse 
% brain connectivity data from 1200 healthy adults.
% It enables the neuroscience 
% research community to discover relationships between brain circuits and 
% individual behaviors. This helps to understand a wide range of brain disorders.
% The HCP project provides database services (ConnectomeDB) for storing and 
% sharing primary and processed data freely, and data analysis pipelines that 
% are available under an open-source license.

The HCP developed a set of pre-processing pipelines to process structural,
functional, and diffusion MRI data acquired in the project. We focus on HCP
pre-processing pipelines for structural data, and particularly
PreFreeSurfer and FreeSurfer. 
A detailed description of the analyses done in these
pipelines is available in~\cite{glasser2013}. 
In summary, PreFreeSurfer consists of the following steps: 
\begin{itemize}
\item Distortion Correction (DC), 
\item Anatomical Average (AAve), 
\item ACPC (Anterior Commissure, Posterior Commissure) Alignment (ACPC-A), 
\item Brain Extraction (BExt), 
\item Bias Field Correction (BFC), 
\item Atlas-Registration (AR).
\end{itemize}
And FreeSurfer consists of the following ones:
\begin{itemize}
\item Downsampling images, 
\item T1w image registration, 
\item Placement of surface images, 
\item Folding-based surface registration.
\end{itemize}
The average processing time per subject is approximately 2 hours for
PreFreeSurfer and 8 hours for FreeSurfer. The average output file size is
approximately 2.7~GB for PreFreeSurfer and 4.1~GB for FreeSurfer.
\tristan{Ali, are these numbers per subject or for 20 subjects?}

We randomly selected 20 unprocessed subjects 
from the HCP data release S500 
available in \href{https://db.humanconnectome.org}{the ConnectomDB repository}. 
\tristan{Can you let me know where the data is located, so that I could describe it here?}
See~\cite{van2013wu} for a discussion of these various modalities.

% HCP data collected from different types of imaging techniques, including 
% structural imaging (sMRI), functional imaging (fMRI), and diffusion imaging (dMRI).
% The structural images include T1-weighted (T1w) and T2-weighted (T2w) images, which 
% help in the diagnosis of brain injury.
% The images are indexed with a suffix if several scans of the same modality were acquired.
% For example, some data may include two T1w images with low and high resolutions.
% The functional images, including task-based fMRI and resting-state scans, 
% enable the measurement of functional activations within brain areas. 
% Diffusion imaging is another kind of MRI technique, which measures 
% the anatomical connectivity between regions.

@@

We built Docker images for the HCP pre-processing pipelines v3.19.0
(PreFreeSurfer and FreeSurfer) in CentOS 6.9 (Final) and CentOS 7.4 (Core),
a widely-used operating system in neurosciences
~\cite{hanke2011neuroscience}. Container images are available on
\href{https://hub.docker.com/r/bigdatalabteam/hcp-prefreesurfer/}{DockerHub}.
They contain the software dependencies of the HCP pipelines, including FSL
(version 5.0.6), FreeSurfer (version 5.3.0-HCP, CentOS4 build), and
Connectome Workbench (version 1.0).

We processed the 20 subjects with PreFreeSurfer and FreeSurfer, using the 2
CentOS versions. Each subject was processed twice on the same operating
system to detect potential within-OS variability, for instance coming from
pseudo-random operations.

\tristan{See what to do with this paragraph}
Within-OS variability may also be coming from other factors than
pseudo-randomness. For instance, the same files may have different metadata
such as timestamps, file paths, and other execution-specific information.
To ensure that the differences are not related to the metadata, we used
specific functions from FreeSurfer tool including \emph{mri\_diff},
\emph{mris\_diff}, and \emph{lta\_diff}. These functions compare files
based on the image dimensions, voxel intensities, the number of vertices,
and geometry information (see~\cite{fischl2012freesurfer} for more
details) \tristan{Ali, are these functions used by \toolname to compare results?}.

In addition, we used the Dice coefficient metric to evaluate the
similarity of image segmentation results, defined by:
\[DICE=\frac{2|X \cap Y|}{|X| + |Y|}\].

% Finally, to cluster the subjects, the threshold value is set to zero in the clustering method. 
% Also, we used the nearest neighbor algorithm to calculate the distance between clusters.

@@

We did not observe any within-OS difference in PreFreeSurfer. In
FreeSurfer, we identified 2 processes leading to within-OS differences due
to the use of pseudo-random numbers: image registration with
\texttt{mri\_segreg}, and cortical surface curvature estimations with
\texttt{mris\_curvature}. To remove these differences, we fixed the random
seed used by FreeSurfer using the \texttt{--seed} command-line option.

@@

We identified four types of subjects with different PreFreeSurfer
provenance graphs (Table~\ref{table:data-clusters}). Differences between
subject types are coming from different  numbers of T1 and T2 images in the
raw data: subjects of type 1 have 2 T1 images and 2 T2 images, while
subjects of type 2 only have 1 T1 image and 1 T2 image. We verified that
the provenance graphs were identical for all subjects of the same type, for
both versions of CentOS.

\begin{table}
\centering
\begin{threeparttable}
\caption{Types of provenance graphs in PreFreeSurfer.}
\label{table:data-clusters}

\begin{tabular}{cccccc}
\toprule
       &                        &  \multicolumn{4}{c}{Image modalities in subject data}    \\ 
\cmidrule(lr){3-6}       
Type   &   \makecell{Number of \\ Subjects}   &  T1w\_1          & T1w\_2      & T2w\_1          & T2w\_2        \\ \midrule
1      &               9                      &   \ding{51}      &   \ding{51} &   \ding{51}     &   \ding{51}   \\ 
2      &               8                      &   \ding{51}      &             &   \ding{51}     &               \\ 
3      &               1                      &   \ding{51}      &             &   \ding{51}     &   \ding{51}   \\ 
4      &               2                      &   \ding{51}      &   \ding{51} &   \ding{51}     &               \\ 
\bottomrule
\end{tabular}
\end{threeparttable}
\end{table}

Figure~\ref{fig:pfs_freq} shows the frequency of non-reproducible pipeline processes
by PreFreeSurfer pipeline step.
Differences were observed in linear registration 
with FSL FLIRT (in ACPC-Alignment, Brain Extraction, Distortion Correction, 
Atlas Registration), in non-linear registration with FSL FNIRT (in Brain Extraction 
and Altas Registration), and in image warping with FSL \texttt{new\_invwarp} (in Brain Extraction 
and Atlas Registration). Differences were also observed in image mean 
computations with FSL maths  (in Anatomical Average). \tristan{figure and caption to be fixed.}

Two types of differences are observed: between-process and between-subject.
Between-process differences occur when the same tool behaves differently in different process instances.
For example, \texttt{flirt}
always produces files without differences in the \texttt{AAve} sub-pipeline but 
almost always creates different files in the \texttt{AAli} sub-pipeline.
Conversely, the same process can be reproducible for some subjects but not for others.

% Between subject differences can be a result of different subject types. 
% For instance, the subjects contain two T1 images need more calculations 
% to produce an average image between them.

% make figure in two columns use {figure*}
\begin{figure*}
\centering
  \includegraphics[width=\textwidth]{images/pfs_heatmap.png}
  \caption{Frequency heatmap of non-reproducible PreFreeSurfer processes (N=20). 
           Rows refer to PreFreeSurfer pipeline steps.
           Cell labels describe the frequency distribution of the instances as ``N [M] x Y" 
           in which N is the proportion of the process 
           that creates differences out of all M times that is occurred. 
           Also, times Y shows the number of processes with the same distribution.
           For example, the cell of FLIRT tool in the sub-pipeline AAve shows 10 instances of the process 
           with the frequency of 11 that introduce no differences.
           The range between red and green colors shows the highest and lowest 
           proportion of occurrences of the processes with differences.
           Also, grey cells indicate that the tool is not executed in the sub-pipeline.}
  \label{fig:pfs_freq}
\end{figure*}

To give an idea of the observed differences, Figure~\ref{fig:fnirt_result} 
compares results of ACPC-Alignment in PreFreeSurfer. \tristan{Ali, there is no
FNIRT process in ACPC-Alignment, please check.}

\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{images/t2w_alignment.png} 
    \caption{Differences between FNIRT results from PreFreeSurfer 
    (checkerboard image \emph{T2w\_acpc\_to\_MNI\_nonli.nii.gz}) (CentOS6 vs. 
    CentOS7). An animated version of the comparison is available 
    \href{https://github.com/ali4006/HCP-reproducibility-paper/blob/master/images/pfs_t2w_alignment.gif}
    {here}.
} 
    \label{fig:fnirt_result}
\end{figure}

@@

The only non-reproducible process identified in FreeSurfer was
\texttt{mris\_make\_surfaces}, which was the case for 10 out of 20
subjects. \tristan{do you know why this was the case: is this CLI
dynamically linked?} This process generates the cortical and white matter
surfaces. However, this does not mean that FreeSurfer results were
identical for the remaining 10 subjects and for the other processes, due to the
propagation of PreFreeSurfer differences.

For instance, Figure~\ref{fig:tissue_class} illustrates the sum of
binarized between-OS differences among the 20 whole-brain segmentations. An
animated comparison of the segmentations obtained in 1 subject is available
\href{https://github.com/ali4006/HCP-reproducibility-paper/blob/master/images/fs_brain_segmentation.gif}
{here}.

The Dice coefficients associated with the 44 regions segmented by
FreeSurfer are shown in Figure~\ref{fig:scatter_plot}. This figure shows
high Dice coefficients (more than 0.95) in large regions and very low Dice
coefficients (below 0.5) in small structures particularly in the vessels
and lateral ventricles.

\begin{figure}
%  \includegraphics{brain\_classification}
\centering
  \includegraphics[width=\columnwidth]{images/brain_segmentation_mni.png} 
  \caption{Sum of binarized between-OS differences between whole-brain FreeSurfer segmentations (N=20),
           resampled and overlaid to the MNI152 volume template.
} 
  \label{fig:tissue_class}
\end{figure}

@@

  The similar findings in another study~\cite{bowring2019exploring} were reported 
  the effect of between-subject differences on reproducibility.
